Elephant Seal Age JAE Intraspecific Variation

### 2024-05-22 paper/paperv3.qmd
Tried filling "gaps", but will use mark-recap instead

```{r}
#| label: setup
#| include: false  
library(ggthemes)
library(cowplot)
library(lme4)
library(lmerTest) # Loading this library will automatically append p-values to lmer models estimated using something called Satterthwaite's method
library(merTools)
library(tidyverse)
library(ggeffects)
library(patchwork)
library(segmented)
source(here::here("R/format.R"))

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = age / 10)

@@

#

@@

We predicted that breeding success would decline with age. We found that breeding success increases to prime age, then decreases at older ages (@fig-age_dist).

@@

Trying segmented analysis. Note: `segmented` does NOT support mixed effect GLMs!

```{r}
#| label: age_breeding_seg

breeding_mod2 <- glm(
  observed_int ~ age,
  data = sealdat, 
  family = "binomial"
)
breed_nseg <- lapply(
  list(15, c(9, 15)),
  \(.psi) {
    segmented(breeding_mod2, seg.Z = ~age, psi = .psi)
  }
)
AIC(breeding_mod2, breed_nseg[[1]], breed_nseg[[2]]) %>% 
  as_tibble() %>% 
  mutate(model = paste(1:3, "segments")) %>% 
  transmute(model, AIC, dAIC = AIC - min(AIC)) %>% 
  arrange(dAIC)

breeding_best <- breed_nseg[[1]]
breeding_pred <- tibble(age = seq(4, 22, length.out = 100)) %>% 
  predict(breeding_best, 
          newdata = ., 
          type = "response", 
          interval = "confidence") %>% 
  as_tibble() %>% 
  set_names(c("bred", "bred_lwr", "bred_upr")) %>% 
  mutate(age = seq(4, 22, length.out = 100))
breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = sum(!is.na(observed)), 
            perc_breed = sum(observed == "Breeder") / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age < 22)

ggplot(breeding_pred, aes(age)) +
  geom_ribbon(aes(ymin = bred_lwr, ymax = bred_upr),
              fill = "grey90") + 
  geom_line(aes(y = bred)) +
  geom_point(aes(y = perc_breed), breed_nonbreed) + 
  geom_errorbar(aes(ymin = perc_breed - se, ymax = perc_breed + se), 
                breed_nonbreed,
                width = 0) +
  theme_classic()

@@

age_dist <- sealdat %>% 
  filter(age < 22) %>% 
  group_by(age) %>% 
  count(age) %>% 
  ggplot(aes(x = age, y = n)) + 
  geom_col(fill = "#829BD1") + 
  labs(x = "Age (Years)", y = "Number of individuals") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))

>>>>

```{r}
#| label: setup
#| include: false  
library(ggthemes)
library(cowplot)
library(lme4)
library(lmerTest) # Loading this library will automatically append p-values to lmer models estimated using something called Satterthwaite's method
library(merTools)
library(tidyverse)
library(ggeffects)
library(ggridges)
library(patchwork)
source(here::here("R/format.R"))

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = age / 10) %>% 
  group_by(animalID) %>% 
  mutate(longevity = ifelse(max(year) < 2020,
                            max(age),
                            NA))

all_years <- sealdat %>% 
  group_by(animalID, yearborn, longevity) %>% 
  summarize(age4 = yearborn[1] + 4,
            first_obs = min(year),
            last_obs = max(year),
            .groups = "keep") %>% 
  reframe(year = seq(age4, last_obs)) %>% 
  filter(between(year, 1987, 2020)) %>% 
  left_join(select(sealdat, animalID, year, observed),
            by = c("animalID", "year")) %>% 
  mutate(observed2 = factor(ifelse(is.na(observed), 
                                   "Non-breeder", 
                                   observed),
                            levels = c("Non-breeder", "Breeder")),
         year_fct = factor(year),
         age = year - yearborn,
         age10 = age / 10)

@@

@@

~~We predicted that breeding success would decline with age. We found that breeding success increases to prime age, then decreases at older ages (@fig-age_dist).~~

```{r}
#| label: breeding-mod 
#| fig-width: 9
#| fig-height: 6

breed_age_mod <- glmer(
  observed2 ~ age10 + (1 | animalID) + (1 | year_fct),
  data = all_years, 
  family = "binomial",
  control = glmerControl(optimizer="bobyqa")
)

year_ranef <- ranef(breed_age_mod)$year_fct %>% 
  as_tibble(rownames = "year") %>% 
  mutate(year = as.numeric(year)) %>% 
  rename(ranef = 2) %>% 
  arrange(ranef)

year_ranef2 <- slice(year_ranef, 
                     round(seq(1, nrow(year_ranef), length.out = 5)))

indiv_ranef <- ranef(breed_age_mod)$animalID %>% 
  as_tibble(rownames = "animalID") %>% 
  mutate(animalID = factor(animalID)) %>% 
  rename(ranef = 2) %>% 
  arrange(ranef)

indiv_ranef2 <- slice(indiv_ranef, 
                      round(seq(1, nrow(indiv_ranef), length.out = 5)))

breed_age_pred_year <- ggpredict(breed_age_mod, 
                                 terms = c("age10 [all]", "year_fct"), 
                                 type = "random",
                                 interval = "confidence") %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         year_fct = group,
         year = as.numeric(as.character(group))) %>% 
  semi_join(year_ranef2, by = "year") %>% 
  filter(age <= 20)

breed_age_pred_indiv <- ggpredict(breed_age_mod, 
                                  terms = c("age10 [all]", "animalID"), 
                                  type = "random",
                                  interval = "confidence") %>% 
  as_tibble() %>% 
  mutate(age = x * 10,
         animalID = group) %>% 
  semi_join(indiv_ranef2, by = "animalID") %>% 
  filter(age <= 20)

breed_age_pred <- ggpredict(breed_age_mod, 
                                 terms = "age10 [all]", 
                                 type = "fixed",
                                 interval = "confidence") %>% 
  as_tibble() %>% 
  mutate(age = x * 10) %>% 
  filter(age <= 20)

breed_nonbreed <- all_years %>% 
  group_by(age) %>% 
  summarize(n_observed = sum(!is.na(observed2)), 
            perc_breed = sum(observed2 == "Breeder") / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age <= 20)

pred_year_plot <- ggplot(breed_age_pred_year, aes(age)) +
  # Prediction across years (fixed effect only)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              breed_age_pred,
              fill = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted), 
            breed_age_pred,
            color = "firebrick", linewidth = 1.5) +
  # Predictions for specific years (random effect)
  geom_line(aes(y = predicted, group = year_fct), 
            color = "cornflowerblue", linewidth = 1) +
  # Raw data
  geom_pointrange(aes(y = perc_breed, 
                      ymin = perc_breed - se, 
                      ymax = perc_breed + se),
                  breed_nonbreed) + 
  # Year labels
  geom_text(aes(y = predicted, label = year_fct), 
            filter(breed_age_pred_year, age == 20),
            x = 20.2,
            hjust = 0,
            size = 3) +
  scale_x_continuous("Age", limits = c(NA, 21)) +
  scale_y_continuous("Breeding %", 
                     limits = c(0, 1), 
                     labels = scales::percent) +
  theme_classic()

pred_indiv_plot <- ggplot(breed_age_pred_indiv, aes(age)) +
  # Prediction across years (fixed effect only)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              breed_age_pred,
              fill = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted), 
            breed_age_pred,
            color = "firebrick", linewidth = 1.5) +
  # Predictions for specific years (random effect)
  geom_line(aes(y = predicted, group = animalID), 
            color = "cornflowerblue", linewidth = 1) +
  # Raw data
  geom_pointrange(aes(y = perc_breed, 
                      ymin = perc_breed - se, 
                      ymax = perc_breed + se),
                  breed_nonbreed) + 
  # Individual labels
  geom_text(aes(y = predicted, label = animalID), 
            filter(breed_age_pred_indiv, age == 20),
            x = 20.2,
            hjust = 0,
            size = 3) +
  scale_x_continuous("Age", limits = c(NA, 21)) +
  scale_y_continuous("Breeding %", 
                     limits = c(0, 1), 
                     labels = scales::percent) +
  theme_classic()

pred_year_plot | pred_indiv_plot

year_plot <- all_years %>% 
  filter(age <= 20) %>% 
  left_join(year_ranef, by = "year") %>% 
  group_by(age) %>% 
  summarize(across(ranef, list(mean = mean, sd = sd))) %>% 
  ggplot(aes(age, ranef_mean)) +
  geom_pointrange(aes(ymin = ranef_mean - ranef_sd,
                      ymax = ranef_mean + ranef_sd)) +
  geom_hline(yintercept = 0, color = "firebrick", linewidth = 1.5) +
  labs(x = "Age", y = "Year ranef") +
  theme_classic()

indiv_plot <- all_years %>% 
  filter(age <= 20) %>% 
  left_join(indiv_ranef, by = "animalID") %>% 
  group_by(age) %>% 
  summarize(across(ranef, list(mean = mean, sd = sd))) %>% 
  ggplot(aes(age, ranef_mean)) +
  geom_pointrange(aes(ymin = ranef_mean - ranef_sd,
                      ymax = ranef_mean + ranef_sd)) +
  geom_hline(yintercept = 0, color = "firebrick", linewidth = 1.5) +
  labs(x = "Age", y = "Indiv ranef") +
  theme_classic()

year_plot | indiv_plot

```{r}
#| label: age_breeding_b4after

senescence_age <- 12

breeding_mod_b4 <- glmer(
  observed2 ~ age10 + (1 | animalID) + (1 | year_fct),
  data = filter(all_years, age <= senescence_age), 
  family = "binomial",
  control = glmerControl(optimizer="bobyqa")
)

breeding_mod_after <- glmer(
  observed2 ~ age10 + (1 | animalID) + (1 | year_fct),
  data = filter(all_years, age > senescence_age), 
  family = "binomial",
  control = glmerControl(optimizer="bobyqa")
)

summary(breeding_mod_b4)
summary(breeding_mod_after)

```{r}
#| label: fig-age_breeding
#| fig-cap: "A: Age structure of northern elephant seals at Año Nuevo. B: Percentage of each age class that were observed breeding."
#| fig-width: 9

age_dist <- all_years %>% 
  filter(age < 22) %>% 
  group_by(age) %>% 
  count(age) %>% 
  ggplot(aes(x = age, y = n)) + 
  geom_col(fill = "#829BD1") + 
  labs(x = "Age (Years)", y = "Number of individuals") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))

breed_pred <- rbind(
  ggpredict(breeding_mod_b4,
            terms = "age10 [all]") %>% 
    as_tibble(),
  ggpredict(breeding_mod_after,
            terms = "age10 [all]") %>% 
    as_tibble()
) %>% 
    mutate(age = x * 10,
           age_class = ifelse(age <= senescence_age,
                              "Before",
                              "After"))

breed_nonbreed <- all_years %>% 
  group_by(age) %>% 
  summarize(n_observed = sum(!is.na(observed2)), 
            perc_breed = sum(observed2 == "Breeder") / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age < 22) %>%
  ggplot(aes(x = age, y = perc_breed)) + 
  geom_ribbon(aes(age, ymin = conf.low, ymax = conf.high, group = age_class),
              breed_pred,
              inherit.aes = FALSE,
              fill = "#829BD1",
              alpha = 0.25) +
  geom_line(aes(age, predicted, group = age_class),
            breed_pred,
            inherit.aes = FALSE,
            color = "#829BD1",
            linewidth = 1) +
  geom_point() + 
  geom_errorbar(aes(ymin = perc_breed - se, ymax = perc_breed + se), 
                width = 0) +
  scale_y_continuous(labels = scales::percent, limits = c(0.4, 1)) +
  expand_limits(x = 23) +
  labs(x = "Maternal Age (Years)", y = "% Breeders (Mean ± SE)") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))

@@

@@

age_dist <- all_years %>% 
  filter(age < 22) %>% 
  group_by(age) %>% 
  count(age) %>% 
  ggplot(aes(x = age, y = n)) + 
  geom_col(fill = "#829BD1") + 
  labs(x = "Age (Years)", y = "Number of individuals") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))
